---
title: "Exploratory data analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      message = FALSE, comment = '',
                      engine.opts = list(bash = "-l"))
```

## Sample:

- 445 individuals from The 1000 Genomes Project (358 from 4 European populations + 87 Yorubans) with RNA-seq available from the Geuvadis consortium.



## 1. VCF processing

### Subset and filtering

Here we select the GEUVADIS individuals, and exclude variants with missing data or which are monomorphic in this subset of individuals.

Next we call an R script that converts VCF to GDS.


```{bash, eval=FALSE, code = readLines("launch_processVCF.sh")}

```

The content of the vcf2gds.R script is the following:

```{r, eval=FALSE, code = readLines("vcf2gds.R")}
```

## Analyses in R

### Packages:

```{r pkgs}
library(tidyverse)
library(SeqArray) 
library(SeqVarTools) 
library(SNPRelate)
library(GENESIS)
library(Biobase)
library(cowplot)
library(ggbeeswarm)
```


### Concatenate GDS files for all chromosomes

```{r VcfToGds, eval=FALSE}
gds_file <- "/home/vitor/heritability/data/kgp/allchrs.gds"

gds_list <- sprintf("/home/vitor/heritability/data/kgp/chr%d.gds", 1:22)

seqMerge(gds_list, gds_file)
```

### Pruning
```{r, eval=FALSE}
# this step takes hours
gds <- seqOpen(gds_file)

set.seed(100)
pruned <- snpgdsLDpruning(gds, 
                          method = "corr", 
                          ld.threshold = sqrt(0.1))

prunedsnps <- unlist(pruned, use.names = FALSE)

seqSetFilter(gds, variant.id = prunedsnps)
seqExport(gds, "/home/vitor/heritability/data/kgp/allchrs_pruned.gds")
seqClose(gds)
```


### PCA on genotype data

```{r}
# Close GDS and open the pruned one
gdsfmt::showfile.gds(closeall = TRUE)

gds <- seqOpen("/home/vitor/heritability/data/kgp/allchrs_pruned.gds")

pca <- snpgdsPCA(gds, num.thread = 16L)

pops <- read_tsv("/raid/genevol/heritability/hla_expression.tsv") %>%
    distinct(subject_id, pop) %>%
    rename(sampleid = subject_id)

pcadf <- as.data.frame(pca$eigenvect) %>%
    as_tibble() %>%
    add_column(sampleid = pca$sample.id, .before = 1) %>%
    inner_join(pops, ., by = "sampleid")
```

```{r pcaplot}
p1 <- ggplot(pcadf, aes(V1, V2, color = pop)) +
    geom_point() +
    scale_color_viridis_d() +
    theme_bw()

p2 <- ggplot(pcadf, aes(V2, V3, color = pop)) +
    geom_point() +
    scale_color_viridis_d() +
    theme_bw()

p3 <- ggplot(pcadf, aes(V3, V4, color = pop)) +
    geom_point() +
    scale_color_viridis_d() +
    theme_bw()

p_legend <- get_legend(p1)

plot_grid(p1 + theme(legend.position = "none"), 
          p_legend,
          p2 + theme(legend.position = "none"), 
          p3 + theme(legend.position = "none"),
          nrow = 2)
```

### Phenotypes (HLA-A expression in Geuvadis)

```{r}
hla_expression <- read_tsv("/raid/genevol/heritability/hla_expression.tsv") %>%
    filter(gene_name == "HLA-A", subject_id %in% seqGetData(gds, "sample.id")) 
```

### Expression across populations

```{r}
ggplot(hla_expression, aes(reorder(pop, tpm), tpm)) +
    geom_quasirandom(aes(color = pop), method = "smiley") +
    geom_boxplot(fill = NA, outlier.color = NA) +
    scale_color_viridis_d() +
    theme_bw() +
    labs(x = NULL, y = "TPM")
```

### Expression across laboratories

```{r}
ggplot(hla_expression, aes(reorder(lab, tpm), tpm)) +
    geom_quasirandom(aes(color = lab), method = "smiley") +
    geom_boxplot(fill = NA, outlier.color = NA) +
    theme_bw() +
    labs(x = NULL, y = "TPM")
```

### Expression between males and females

```{r}
ggplot(hla_expression, aes(reorder(sex, tpm), tpm)) +
    geom_quasirandom(aes(color = sex), method = "smiley") +
    geom_boxplot(fill = NA, outlier.color = NA) +
    theme_bw() +
    labs(x = NULL, y = "TPM")
```

Here we got to evaluate whether such differences across groupings are adequate for an analysis with the combined dataset, or whether they can be accounted for by using covariates.

Before we continue, we need to transform our phenotype data.frame into an "Annotated data frame", which is a data structure provided by `Biobase`. This means that we have to add metadata to our `hla_expression` data.frame, and include a `sample.id` column as a requirement for `GENESIS`. Since we do not have multiple samples for the same individuals, we will just create the `sample.id` column as a copy of `subject_id`.


```{r}
metadata <- 
    data.frame(labelDescription = c("sample identifier",
                                    "subject identifier",
                                    "laboratory of RNA sequencing",
                                    "subject's sex",
                                    "subject's population",
                                    "PC 1",
                                    "PC 2",
                                    "PC 3",
                                    "expression levels in TPM"))

annotphen <- hla_expression %>%
    mutate(sample.id = subject_id) %>%
    left_join(select(pcadf, sampleid, V1:V3), 
              by = c("subject_id" = "sampleid")) %>%
    select(sample.id, subject_id, lab, sex, pop, V1, V2, V3, tpm) %>%
    as.data.frame() %>%
    AnnotatedDataFrame(metadata)

# access the metadata with the varMetadata() function
varMetadata(annotphen)

# access the data with the pData() function
head(pData(annotphen))
```


## GRM

### GCTA

We will use the `SNPRelate` package to compute a GRM. We will begin with the GCTA method.

```{r}
# Close GDS and open the pruned one
gdsfmt::showfile.gds(closeall = TRUE)

pruned <- seqOpen("/home/vitor/heritability/data/kgp/allchrs_pruned.gds")

# Computar a GRM
grm_obj <- snpgdsGRM(pruned, method = "GCTA", num.thread = 16L)
```

We extract and rename the matrix
```{r}
sample_ids <- seqGetData(pruned, "sample.id")

grm <- grm_obj$grm
rownames(grm) <- sample_ids
colnames(grm) <- sample_ids

# first 5 individuals:
grm[1:5, 1:5]
```

## Fitting the Null model

The first step in finding genetic variants which are associated with a phenotype is preparing null model.

We fit a null model which adjusts gene expression according to covariates such as population of origin, laboratory of sequencing, and sex. We also account for the relatedness between individuals (GRM).


```{r eval=FALSE}
mod_null <- fitNullModel(annotphen, 
                         outcome = "tpm", 
                         covars = c("lab", "sex", "pop", "V1", "V2", "V3"),
                         cov.mat = grm)

#Error in chol.default(Sigma) : the leading minor of order 445 is not positive definite
#12. chol.default(Sigma)
#11. chol(Sigma)
#10. chol(Sigma)
#9. .computeSigmaQuantities(varComp = sigma2.k, covMatList = covMatList, group.idx = group.idx)
#8. .runAIREMLgaussian(y, X, start = start, covMatList = covMatList, group.idx = group.idx, AIREML.tol = AIREML.tol, drop.zeros = drop.zeros, max.iter = max.iter, EM.iter = EM.iter, verbose = verbose)
#7. .fitNullModel(y = desmat$y, X = desmat$X, covMatList = cov.mat, group.idx = desmat$group.idx, family = family, start = start, AIREML.tol = AIREML.tol, max.iter = max.iter, EM.iter = EM.iter, drop.zeros = drop.zeros, return.small = return.small, verbose = verbose)
#6. .local(x, ...)
#5. fitNullModel(x, outcome, covars, cov.mat, group.var, ...)
#4. fitNullModel(x, outcome, covars, cov.mat, group.var, ...)
#3. .local(x, ...)
#2. fitNullModel(annotphen, outcome = "tpm", covars = c("lab", "sex", "pop", "V1", "V2", "V3"), cov.mat = grm)
#1. fitNullModel(annotphen, outcome = "tpm", covars = c("lab", "sex", "pop", "V1", "V2", "V3"), cov.mat = grm)
```

Now that we have a Null model adjusting expression levels for population, sex, laboratory, population genetic structure, and the relatedness, we can test for the association of the genetic variants with expression levels.

The first step is to create a `SeqVarData` object including both the GDS (genotypes) and the Annotated data.frame (phenotypes). 

Then we will use the `assocTestSingle` function to assess the effect of each variant.

```{r eval=FALSE}
# order individuals according to GDS
pData(annotphen) <- pData(annotphen) %>%
    mutate(subject_id = factor(subject_id, levels = seqGetData(pruned, "sample.id"))) %>%
    arrange(subject_id)

# create SeqVarData object and iterator
seqData <- SeqVarData(pruned, sampleData = annotphen)
iterator <- SeqVarBlockIterator(seqData, verbose = FALSE)

# test
assoc <- assocTestSingle(iterator, mod_null)

head(assoc)
```


`GENESIS` includes the `varCompCI` to compute the proportion of variance explained (heritability) by each random effect, with a 95% CI: 

```{r eval=FALSE}
varCompCI(nullmod, prop = TRUE)
```


## Packages version

```{r echo=FALSE}
devtools::session_info()
```

